---
title: "Reproducible Research: Peer Assessment 1"
author: "Yazad Jal"
output: 
  html_document: 
    keep_md: yes
keep_md: true
---

## Loading and preprocessing the data

We first download the zipfile from the web and save it in the working directory

```{r file_download, message=FALSE, warning=FALSE}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- paste0(getwd(),"/","activity.zip")
download.file(fileurl, destfile, method = "curl", quiet = TRUE)
```

Then we unzip the downloaded file and get a feel for the data using str

```{r data_feel}
unzip("activity.zip")
activity <- read.csv("activity.csv")
str(activity)
```

To calculate total number steps per day, I created a new data frame called totalsteps and then made a histogram. 

```{r totalsteps}
totalsteps<-aggregate(steps~date,data=activity,sum,na.rm=TRUE)
hist(totalsteps$steps, 
        col = "royalblue", border = "royalblue", density = 50,
        xlab = "Number of Steps", main = "Total steps per day",
        ylim = c(0,30))
```

## What is mean total number of steps taken per day?

Calculating the mean and median of the total number of steps taken per day:

```{r mean_median}
mn <- mean(totalsteps$steps)
md <- median(totalsteps$steps)
```

```{r, echo=FALSE}
mnr <- round(mn, digits = 2)
```

The mean total number of steps per day is `r as.character(mnr)` and the median is `r md`


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
