---
title: "Reproducible Research: Peer Assessment 1"
author: "Yazad Jal"
output: 
  html_document: 
    keep_md: yes
keep_md: true
---

## Loading and preprocessing the data

We first download the zipfile from the web and save it in the working directory

```{r file_download, message=FALSE, warning=FALSE}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- paste0(getwd(),"/","activity.zip")
download.file(fileurl, destfile, method = "curl", quiet = TRUE)
```

Then we unzip the downloaded file and get a feel for the data using str

```{r data_feel}
unzip("activity.zip")
activity <- read.csv("activity.csv")
str(activity)
```


To calculate total number steps per day, I created a new data frame called totalsteps and then made a histogram. 

```{r totalsteps}
totalsteps<-aggregate(steps~date,data=activity,sum,na.rm=TRUE)
par(mfrow=c(1,1))
hist(totalsteps$steps, 
        col = "royalblue", border = "royalblue", density = 50,
        xlab = "Number of Steps", main = "Total steps per day",
        ylim = c(0,30))
```


## What is mean total number of steps taken per day?

Calculating the mean and median of the total number of steps taken per day:

```{r mean_median}
mn <- mean(totalsteps$steps)
md <- median(totalsteps$steps)
```

```{r mnr, echo=FALSE}
mnr <- round(mn, digits = 2)
```

The mean total number of steps per day is `r as.character(mnr)` and the median is `r md`


## What is the average daily activity pattern?

First I created a new data frame, fivemin, which aggregates the activity on each five minute interval across the 61 days, and then I plot it, giving us the average daily activity pattern.

```{r fivemin}
fivemin <- aggregate(steps~interval, data = activity, mean, na.rm=TRUE)
plot(fivemin$interval, fivemin$steps, type = "l", col = "royalblue", lwd=3,
     xlab="Time of Day", ylab="Average Number of Steps", ylim=c(0, 250),
     main="Average Daily Activity Pattern", xaxt="n")
axis(side=1, at = c(0, 400, 800, 1200, 1600, 2000, 2400), 
     labels = c("midnight", "4am", "8am", "12pm", "4pm", "8pm", "midnight"))
```


Computing the five minute interval with the maximum number of steps:
```{r max_fivemin}
max <- fivemin[which.max(fivemin$steps),]$interval
max2 <-paste0(0, max)
max3 <-paste0(0, (max+5))
```

From `r max2` to `r max3` is when the maximum exercise happens.


## Imputing missing values

Calculating and reporting the total number of missing values in the dataset (i.e. the total number of rows with NAs): 
```{r isna}
isna <- sum(is.na(activity$steps))
```

There are `r isna` rows with missing values in the dataset

I've used two strategies to fill in the missing values. The first is using the impute function from the Hmisc package, and creating a new dataset, totalsteps2, and a new histogram. 

```{r Hmisc_method, message=FALSE, warning=FALSE}
library(Hmisc, quietly = TRUE)
activityImputed <- activity
activityImputed$steps <- impute(activity$steps, fun=mean)
totalsteps2<-aggregate(steps~date,data=activityImputed,sum,na.rm=TRUE)
```

```{r Hmisc_impute_hist}
hist(totalsteps2$steps, 
        col = "red", border = "red", density = 50,
        xlab = "Number of Steps", main = "Total steps per day",
        ylim = c(0,35))
```

The new mean and median using this method are:
```{r Hmisc_impute_newmean}
mean(totalsteps2$steps)
median(totalsteps2$steps)
```

The mean has statyed the same (it should, as the method imputed the mean and replaced the NA values with it). The median has changed and is now equal to the mean. 

Comparing the histograms:
```{r comparing_hist1}
par(mfrow=c(1,2))
hist(totalsteps$steps, 
        col = "royalblue", border = "royalblue", density = 50,
        xlab = "Number of Steps", main = "Total steps per day (original)",
        ylim = c(0,35))

hist(totalsteps2$steps, 
        col = "red", border = "red", density = 50,
        xlab = "Number of Steps", main = "Total steps per day (new)",
        ylim = c(0,35))
```

The impact of imputing missing data seems to be to increase the average frequency. 


The second method used is k nearest neighbors from the VIM package.

```{r knn_impute_method, message=FALSE, warning=FALSE, cache=TRUE}
library(VIM)
act <- activity
knnact <- kNN(act)
totalsteps3<-aggregate(steps~date,data=knnact,sum,na.rm=TRUE)
```

The new mean and median using this method are:
```{r knn_impute_newmean}
mean(totalsteps3$steps)
median(totalsteps3$steps)
```

The mean and median have both decreased. That is because the method has imputed 0 values for many of the NA values.

```{r knn_impute_hist}
par(mfrow=c(1,1))
hist(totalsteps2$steps, 
        col = "seagreen3", border = "seagreen3", density = 50,
        xlab = "Number of Steps", main = "Total steps per day",
        ylim = c(0,35))
```


Comparing both methods with the original histogram
```{r comparing_hist2}
par(mfrow=c(1,3))
hist(totalsteps$steps, 
        col = "royalblue", border = "royalblue", density = 50,
        xlab = "Number of Steps", main = "Total steps per day (original)",
        ylim = c(0,40))

hist(totalsteps2$steps, 
        col = "red", border = "red", density = 50,
        xlab = "Number of Steps", main = "Total steps per day (hmisc)",
        ylim = c(0,40))

hist(totalsteps2$steps, 
        col = "seagreen3", border = "seagreen3", density = 50,
        xlab = "Number of Steps", main = "Total steps per day (knn)",
        ylim = c(0,40))
```

The impact of imputing missing data seems to increase the average frequency, in both methods.


## Are there differences in activity patterns between weekdays and weekends?

To answer this, I've used the original data frame, activity, (with the NAs) and added a new column denoting weekday / weekend. I then use the filter function from the dplyr package to filter for weekday / weekend and aggregated over the steps in each five minute interval

```{r weekday_weekend, message=FALSE, warning=FALSE}
activity$day <- ifelse(as.POSIXlt(as.Date(activity$date))$wday%%6 == 0,
                       "weekend", "weekday")
library(dplyr)
weekdays <- filter(activity, day == "weekday")
weekends <- filter(activity, day == "weekend")
fivemin_wd <- aggregate(steps~interval, data = weekdays, mean, na.rm=TRUE)
fivemin_we <- aggregate(steps~interval, data = weekends, mean, na.rm=TRUE)
```

Plotting the weekday and weekend activity patterns in a panel plot:

```{r weekday_weekend_plot}
par(mfrow=c(1,2))
plot(fivemin_wd$interval, fivemin_wd$steps, type = "l", 
     col = "royalblue", lwd=3, ylim=c(0, 250),
     xlab="Time of Day", ylab="Average Number of Steps", 
     main="Weekday Activity Pattern", xaxt="n")
axis(side=1, at = c(0, 400, 800, 1200, 1600, 2000, 2400), 
     labels = c("midnight", "4am", "8am", "12pm", "4pm", "8pm", "midnight"))
plot(fivemin_we$interval, fivemin_we$steps, type = "l", 
     col = "red", lwd=3,ylim=c(0, 250),
     xlab="Time of Day", ylab="Average Number of Steps", 
     main="Weekend Activity Pattern", xaxt="n")
axis(side=1, at = c(0, 400, 800, 1200, 1600, 2000, 2400), 
     labels = c("midnight", "4am", "8am", "12pm", "4pm", "8pm", "midnight"))
```

The weekend pattern is significantly different from the weekday pattern. Weekends do not have a unique peak and there are generally fewer steps walked during weekends.